Sampling requirements to detect ecosystem change

### 2020-03-25 R/manuscript_draft.Rmd
updated code to build theoretical plots

Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, https://orcid.org/0000-0001-8351-7539 $^{2}$Department of Biology, University of Vermont, VT 05405, USA, https://orcid.org/0000-0002-0768-9555
* Both authors contributed contributed equally and order was chosen alphabetically / by latitudinal gradient of birth place

@@

Submission as: Article at \emph{Ecological Applications}

@@

\textsc{Rosalie Bruel$^{1}$\footnote{The authors contributed equally to this work.} and Easton R. White$^{2}$}
\vspace{3 mm}

@@

# Sigma results
sigma_vec = seq(0.1,30,4)
min_time  = rep(0,times=length(sigma_vec))
sigma_output <- data.frame(sigma_vec,min_time)

@@

# Melt the 3 dataset to be able to use facet_wrap()
colnames(sigma) <- colnames(phi) <- colnames(shift_size) <- c("X", "Param", "min_time")
simulation_all <- 
  rbind(cbind("Parameter" = rep("Standard deviation", nrow(sigma)), sigma),
      cbind("Parameter" = rep("Autocorrelation at lag-1", nrow(phi)), phi),
      cbind("Parameter" = rep("Normalized amplitude of shift", nrow(shift_size)), shift_size))
# param_vr <- data.frame("Parameter" = unique(simulation_all$Parameter),
#                        "value" = c(ar_vr$acf[2], sd_vr, shift_vr))

p1 <- 
ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[1]), aes(Param, min_time)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=ar_vr$acf[2], lty=2) + 
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('a. {Parameter}')) +
  ylim(20,50) +
  labs(x=c("sigma"), y= "Minimum number of samples required") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))
p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = sd_vr, lty=2) +
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('b. {Parameter}')) +
  ylim(20,50) +
  labs(x="phi", y= "") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))
p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = shift_vr, lty=2) +
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('c. {Parameter}')) +
  ylim(20,50) +
  labs(x="shift size", y= "") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))

@@

In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size. Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:simulation_min_time}). For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr$acf[2],3)`, the variance is `r round(sd_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.

>>>>

Addresses: $^{1}$Rubenstein Ecosystem Science Laboratory, University of Vermont, VT 05405, USA, https://orcid.org/0000-0001-8351-7539 $^{2}$Department of Biology, University of Vermont, VT 05405, USA, https://orcid.org/0000-0002-0768-9555
* Both authors contributed contributed equally and order was chosen alphabetically

@@

Submission as: Article at \emph{Journal of Applied Ecology}

@@

\textsc{Rosalie Bruel$^{*1}$\footnote{*The authors contributed equally to this work.} and Easton R. White$^{*2}$}
\vspace{3 mm}

@@

# Sigma results
sigma_vec = seq(0.01,1,0.05)
min_time  = rep(0,times=length(sigma_vec))
sigma_output <- data.frame(sigma_vec,min_time)

@@

# Melt the 3 dataset to be able to use facet_wrap()
colnames(sigma) <- colnames(phi) <- colnames(shift_size) <- c("X", "Param", "min_time")
simulation_all <- 
  rbind(cbind("Parameter" = rep("Standard deviation", nrow(sigma)), sigma),
      cbind("Parameter" = rep("Autocorrelation at lag-1", nrow(phi)), phi),
      cbind("Parameter" = rep("Normalized amplitude of shift", nrow(shift_size)), shift_size))
# param_vr <- data.frame("Parameter" = unique(simulation_all$Parameter),
#                        "value" = c(ar_vr, cv_vr, shift_vr))

p1 <- 
ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[1]), aes(Param, min_time)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=ar_vr, lty=2) + 
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('a. {Parameter}')) +
  ylim(20,50) +
  labs(x=c("sigma"), y= "Minimum number of samples required") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))
p2 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[2]), aes(Param, min_time)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = cv_vr, lty=2) +
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('b. {Parameter}')) +
  ylim(20,50) +
  labs(x="phi", y= "") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))
p3 <- ggplot(simulation_all %>% filter(Parameter == unique(simulation_all$Parameter)[3]), aes(Param, min_time))  +
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = shift_vr, lty=2) +
  facet_wrap(~Parameter, scales="free_x", 
             labeller = label_glue('c. {Parameter}')) +
  ylim(20,50) +
  labs(x="shift size", y= "") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_rect(colour="transparent", fill="transparent"),
        strip.text.x = element_text(angle = 0, hjust = 0))

@@

In line with theory about monitoring, we found that the probability of correcting identifying a changepoint decreased with smaller levels of population variability ($\sigma$) and autocorrelation ($\phi$). We also found that the probability of correct changepoint detection increases with larger shift sizes, which is essentially the effect size. Thus, the number of sample required to obtain high statistical power (above 0.8) increased with larger population variability, lower autocorrelation, and smaller shift sizes (Fig. \ref{fig:simulation_min_time}). For perspective, the autocorrelation at lag-1 in the first DCA component we used in the case study example later is `r round(ar_vr,3)`, the coefficient of variation is `r round(cv_vr,3)`, and the normalized amplitude of the shift was `r round(shift_vr,3)`.
